UMAP reveals cryptic population structure and phenotype heterogeneity in large genomic cohorts

### 2019-06-11 ms/main.tex
minor cleanup and text changes

\begin{figure*}
    \centering
    \begin{subfigure}{0.5\columnwidth}
    \includegraphics[width=\columnwidth]{images/HRS_1000G_NP1_UMAP_PC7_NC2_NN15_MD05_pca_hrshisp_added1kgp_2018115153245_1kgp_hisp.pdf}
    \caption{}
    \label{fig:supp_umap_hrs_hisp_1kgp_hisp}
    \end{subfigure}\hfill
        \begin{subfigure}{0.5\columnwidth}
    \includegraphics[width=\columnwidth]{images/Asthma_FullAsian_PC_QC_new_UMAP_PC8_NC2_NN15_MD05_20193117303txt_eth_for_ms_lw2.jpeg}
    \caption{}
        \label{fig:umap_ukbb_indian}
    \end{subfigure}
    \caption{\textbf{Applying UMAP to subets of data can reveal deep population structure.} \textbf{(a)} UMAP on the top $7$ principal components of the self-identified Hispanic population of the HRS reveals a cluster (highlighted). Colouring the points by birthplace shows they were born almost entirely in the Mountain region of the United States (New Mexico, Arizona, Colorado, Utah, Nevada, Wyoming, Idaho, and Montana). When populations from the 1KGP are projected onto the UMAP embedding they do not map to the cluster. Six 1KGP populations are presented:
    CLM, Colombian in Medellin, Colombia;
    IBS, Iberian in Spain;
    MXL, Mexican in Los Angeles, California;
    PEL, Peruvian;
    PUR, Puerto Rican;
    TSI, Tuscani in Italy. Figure~\ref{fig:supp_umap_hrs_hisp_admix} presents the same projection of individuals from the HRS coloured by estimated admixture proportions. \textbf{(b)} UMAP on the top $8$ principal components of the self-identified Asian populations of the UKBB create clusters. Almost all individuals in the highlighted cluster are Indian individuals born in Kenya.}
    \label{fig:subpops}
\end{figure*}

@@

\adpcomment{changed this a bit. or maybe remove most of it?}
A few individuals cluster with populations different from their label: some Mexican individuals cluster with Spanish and Italian populations; some Puerto Rican individuals cluster with African American and Caribbean populations; and one Gambian individual clusters with the Mende of Sierra Leone. Two populations form multiple clusters: Gujarati Indians in Houston, Texas and Punjabi people in Lahore, Pakistan. This clustering effect is robust to the number of components considered (see figure~\ref{fig:supp_megamontage_pc10_50}). Differentiation in the Gujarati population from substructure has been previously noted\cite{reich2009india}. Recontacting of 23andMe participants associated with the clusters show that one of the cluster is highly enriched in individuals from a group in Western India with shared ancestry and patronym. [D. Poznik, 23andme, personal communication, and \cite{23andme}].

%of individual bearing the same population label: a few Mexican individuals cluster with Spanish and Italian individuals, and a few Puerto Ricans cluster with the African Americans and African Caribbeans, likely resulting from ancestry proportions that differ from the majority. One Gambian-identified individual is present in a cluster that is otherwise entirely Mende people from Sierra Leone. Two populations form multiple clusters: Gujarati Indians in Houston, Texas and Punjabi people in Lahore, Pakistan. This clustering effect is robust to the number of components considered (see figure~\ref{fig:supp_megamontage_pc10_50}). Differentiation in the Gujarati population from substructure has been previously noted\cite{reich2009india}. Recontacting of 23andMe participants associated with the clusters show that one of the cluster is highly enriched in individuals from a group in Western India with shared ancestry and patronym. [D. Poznik, 23andme, personal communication, and \cite{23andme}].

%The genetic continuum of admixed populations
\adpcomment{cleaned up, changed subtitle}
\subsubsection*{Admixed individuals fall along a genetic continuum}
The 1KGP sampled individuals from relatively distinct populations, so the data are more likely to form clusters. Most medical cohorts, however, comprise larger numbers of individuals sampled across extended geographical areas. The HRS contains genotype data of 12,454 Americans from a variety of backgrounds. Using UMAP on the first 10 principal components, we demonstrate projections that present a collection of sub-populations and a continuum of genetic variation.

The HRS forms several large clusters, reflecting both ethnicity (figure~\ref{fig:supp_umap_hrs_eth}) and admixture proportions (figure~\ref{fig:umap_hrs_admix}). Gradients in admixture proportion are visible within the predominantly Hispanic cluster, but not within the predominantly Black cluster, perhaps because the variance in ancestry proportions is greater among Hispanics. The "White Not Hispanic" (WNH) group forms several interconnected clusters, and these do not correspond to broad geographical areas (figure~\ref{fig:supp_hrs_born}). Projecting 1KGP populations onto the embedding reveals substructure within the Hispanic cluster, groupings of Finnish individuals within the WNH groups, as well as Italian and Spanish individuals grouping near the White Hispanic population (figure~\ref{fig:supp_hrs_1kgp_projected}). One group of WNH individuals regularly appears at the periphery of the main cluster and does not cluster with any 1KGP populations.

@@

Figure~\ref{fig:umap_ukbb_geo_osgb} presents the UMAP projection from figure~\ref{fig:umap_ukbb_pc10} coloured instead by geographical coordinates from the Ordnance Survey National Grid (OSGB1936), with distances defined as a north or east position relative to the Isles of Scilly. Geographic clusters form in the large White British grouping, reflecting the relationship between genetic and geographic distance, as has been observed in Europe-wide data\cite{novembre2008europe}. Most admixture lines have coordinates in the south east of the UK, where the city of London is located, likely reflecting its high migrant population. The detailed shape of extended clusters is not stable as we vary the number of PCs included, but the patterns mentioned above are preserved. Figures~\ref{fig:supp_montage_ukbb_eth}, \ref{fig:supp_montage_ukbb_ns}, and~\ref{fig:supp_montage_ukbb_ew} show UMAP plots using the top $40$ PCs from the UKBB. %The shape of the White British cluster is different, and we observe finer patterns of geographic variation, yet the observations made earlier are maintained.

@@

%These results merit further investigation and underscore the exploratory value of UMAP --- PCA alone compresses too many populations together to see finer details (figures~\ref{fig:supp_ukbb_pca_height_res_f}, \ref{fig:supp_ukbb_pca_fev_f}), while t-SNE's clustering breaks the desirable continuity between and within populations~(figures~\ref{fig:supp_ukbb_tsne_height_res_f}, \ref{fig:supp_ukbb_tsne_fev_f}).

@@

%To compare UMAP to t-SNE and assess the role of convergence, we ran UMAP, Python's sklearn implementation of t-SNE, and a hybrid approach where the early exaggeration phase of t-SNE (typically the first 250 iterations) has been replaced by a full UMAP run (600 epochs for the 1KGP data and 230 epochs for the HRS and UKBB data). The three approaches produce similar results in smaller datasets such as the 1KGP and HRS, with the main differences being in run-time and some qualitative factors such as clusters being more interconnected when using UMAP (figures~\ref{fig:supp_tsne_umap_compare_1kgp},~\ref{fig:supp_tsne_umap_compare_hrs}). When applied to the UKBB, UMAP provides a more compelling visualization of the data compared to both versions of t-SNE. Interestingly, the hybrid approach produces a better value for the t-SNE metric than the default t-SNE approach (figure~\ref{fig:supp_tsne_umap_compare_ukbb_graph}), suggesting that UMAP can help standard t-SNE implementation to reach the t-SNE objective. The standard t-SNE initialization generated a large number of arbitrary clusters, while initializing t-SNE with UMAP preserved much of the large scale structure created by the UMAP embedding but lost much fine-scale resolution and the relationship among clusters (figure~\ref{fig:supp_tsne_umap_compare_ukbb}). An animation of t-SNE iterations starting from the UMAP embedding can be seen in the supplementary files. Given these results, we recommend UMAP over t-SNE for large and diverse genomic datasets.

@@

The HRS contains genotype data of 12,454 American individuals across all 50 states who have provided racial identity (10,434 White, 1,652 Black, 368 Other) as well as whether they identify as Hispanic (1,203 total) and, if so, whether they identify as Mexican-American (705 total)\cite{juster1995overview}. We crossed these three variables to form a composite self-reported ethnicity resulting in 10 categories (e.g. White Hispanic Mexican-American), and considered birth regions based on the 10 census regions and divisions used by the US Census Bureau. Admixture proportions for each individual were estimated in \cite{baharian2016great} by assuming ancestral African, Asian, and European populations using RFMIX \cite{Maples:2013fia}. We have scaled each of the three proportions to values between 0 and 255 (with 100\% corresponding to 255), to colour individual points by their estimated admixture represented by RGB where red, green, and blue respectively correspond to African, European, and Asian/Native American ancestry.

@@

\begin{figure*}
    \centering
    \begin{subfigure}{\textwidth}
    \includegraphics[width=0.8\textwidth]{images/HRS_1000G_NP1_UMAP_PC7_NC2_NN15_MD05_pca_hrshisp_added1kgp_2018115153245_admix.pdf}
    \end{subfigure}
    \caption{UMAP of the first 7 principal components of the Hispanic population of the HRS, coloured by estimated admixture proportions.}
    \label{fig:supp_umap_hrs_hisp_admix}
\end{figure*}

@@

\begin{figure}
\centering
   \includegraphics[width=0.6\linewidth]{images/HRS_1000G_NP1_UMAP_PC10_NC2_NN15_MD05_pca_1kgp_onto_hrs_umap_1kgp_onto_hrs_2018112221116_admix.pdf}
   \caption{UMAP on the first 10 principal components of HRS data. colouring individuals by estimated admixture from three ancestral populations reveals considerable diversity in the Hispanic population. This projection coloured by self-identified race and Hispanic status is presented in figure~\ref{fig:supp_umap_hrs_eth}.}
    \label{fig:umap_hrs_admix}
\end{figure}

@@

\begin{figure}[!htb]
    \centering
    \includegraphics[width=0.95\columnwidth]{images/1KGP_tsne_umap.jpeg}
    \caption{Comparing the visualizations of UMAP, standard t-SNE, and t-SNE initialized with a UMAP projection, on the top 10 principal components of the 1KGP. t-SNE used 5000 iterations.}
    \label{fig:supp_tsne_umap_compare_1kgp}
\end{figure}

>>>>

\begin{figure*}
    \centering
    \begin{subfigure}{0.5\columnwidth}
    \includegraphics[width=\columnwidth]{images/HRS_1000G_NP1_UMAP_PC7_NC2_NN15_MD05_pca_hrshisp_added1kgp_2018115153245_1kgp_hisp.pdf}
    \caption{}
    \label{fig:supp_umap_hrs_hisp_1kgp_hisp}
    \end{subfigure}\hfill
        \begin{subfigure}{0.5\columnwidth}
    \includegraphics[width=\columnwidth]{images/Asthma_FullAsian_PC_QC_new_UMAP_PC8_NC2_NN15_MD05_20193117303txt_eth_for_ms_lw2.jpeg}
    \caption{}
        \label{fig:umap_ukbb_indian}
    \end{subfigure}
    \caption{\textbf{Applying UMAP to subets of data can reveal deep population structure.} \textbf{(a)} UMAP on the top $7$ principal components of the self-identified Hispanic population of the HRS reveals a cluster (highlighted). Colouring the points by birthplace shows they were born almost entirely in the Mountain region (in green) of the United States (New Mexico, Arizona, Colorado, Utah, Nevada, Wyoming, Idaho, and Montana). When populations from the 1KGP are projected onto the UMAP embedding they do not map to the cluster. Six 1KGP populations are presented:
    CLM, Colombian in Medellin, Colombia;
    IBS, Iberian in Spain;
    MXL, Mexican in Los Angeles, California;
    PEL, Peruvian;
    PUR, Puerto Rican;
    TSI, Tuscani in Italy. Figure~\ref{fig:supp_umap_hrs_hisp_admix} presents the same projection of individuals from the HRS coloured by estimated admixture proportions. \textbf{(b)} UMAP on the top $8$ principal components of the self-identified Asian populations of the UKBB create clusters. Almost all individuals in the highlighted cluster are Indian individuals born in Kenya. Colours represent self-identified ethnic background, with those identifying as Indian coloured purple. }
    \label{fig:subpops}
\end{figure*}

@@

A few individuals cluster with populations different from their label: some Mexican individuals cluster with Spanish and Italian populations; some Puerto Rican individuals cluster with African American and Caribbean populations; and one Gambian individual clusters with the Mende of Sierra Leone. Two populations form multiple clusters: Gujarati Indians in Houston, Texas and Punjabi people in Lahore, Pakistan. This clustering effect is robust to the number of components considered (see figure~\ref{fig:supp_megamontage_pc10_50}). Differentiation in the Gujarati population from substructure has been previously noted\cite{reich2009india}. Recontacting of 23andMe participants associated with the clusters show that one of the cluster is highly enriched in individuals from a group in Western India with shared ancestry and patronym. [D. Poznik, 23andme, personal communication, and \cite{23andme}].

\subsubsection*{Admixed individuals fall along a genetic continuum}
The 1KGP sampled individuals from relatively distinct populations, so the data are more likely to form clusters. Most medical cohorts, however, comprise larger numbers of individuals sampled across extended geographical areas. The HRS contains genotype data of 12,454 Americans from a variety of backgrounds. Using UMAP on the first 10 principal components, we demonstrate projections that present a collection of sub-populations and a continuum of genetic variation.

The HRS forms several large clusters, reflecting both ethnicity (figure~\ref{fig:supp_umap_hrs_eth}) and admixture proportions (figure~\ref{fig:umap_hrs_admix}). Gradients in admixture proportion are visible within the predominantly Hispanic cluster, but not within the predominantly Black cluster, perhaps because the variance in ancestry proportions is greater among Hispanics. The "White Not Hispanic" (WNH) group forms several interconnected clusters, and these do not correspond to broad geographical areas (figure~\ref{fig:supp_hrs_born}). By generating the PC axes and UMAP embedding for the HRS data in figure~\ref{fig:supp_umap_hrs_eth}, and projecting the 1KGP data onto it, we reveal substructure within the Hispanic cluster, groupings of Finnish individuals within the WNH groups, as well as Italian and Spanish individuals grouping near the White Hispanic population (figure~\ref{fig:supp_hrs_1kgp_projected}). One group of WNH individuals regularly appears at the periphery of the main cluster and does not cluster with any 1KGP populations.

@@

Figure~\ref{fig:umap_ukbb_geo_osgb} presents the UMAP projection from figure~\ref{fig:umap_ukbb_pc10} coloured instead by geographical coordinates from the Ordnance Survey National Grid (OSGB1936), with distances defined as a north or east position relative to the Isles of Scilly. Geographic clusters form in the large White British grouping, reflecting the relationship between genetic and geographic distance, as has been observed in Europe-wide data\cite{novembre2008europe}. Most admixture lines have coordinates in the south east of the UK, where the city of London is located, likely reflecting its high migrant population. The detailed shape of extended clusters is not stable as we vary the number of PCs included, but the patterns mentioned above are preserved. Figures~\ref{fig:supp_montage_ukbb_eth}, \ref{fig:supp_montage_ukbb_ns}, and~\ref{fig:supp_montage_ukbb_ew} show UMAP plots using the top $40$ PCs from the UKBB.

@@

@@

@@

The HRS contains genotype data of 12,454 American individuals across all 50 states who have provided racial identity (10,434 White, 1,652 Black, 368 Other) as well as whether they identify as Hispanic (1,203 total) and, if so, whether they identify as Mexican-American (705 total)\cite{juster1995overview}. We crossed these three variables to form a composite self-reported ethnicity resulting in 10 categories (e.g. White Hispanic Mexican-American), and considered birth regions based on the 10 census regions and divisions used by the US Census Bureau. Admixture proportions for each individual were estimated in \cite{baharian2016great} by assuming ancestral African, Asian, and European populations using RFMIX \cite{Maples:2013fia}. We have scaled each of the three proportions to values between 0 and 255 (with 100\% corresponding to 255), to colour individual points by their estimated admixture represented by RGB where red, green, and blue respectively correspond to African, European, and Asian/Native American ancestry. To project 1KGP data on HRS embeddings, as in figures~\ref{fig:supp_umap_hrs_hisp_1kgp_hisp} and \ref{fig:supp_hrs_1kgp_projected}, we created the PC axes and UMAP embedding for the HRS data and then projected the 1KGP data onto them.

@@

\begin{figure*}
    \centering
    \begin{subfigure}{\textwidth}
    \includegraphics[width=0.8\textwidth]{images/HRS_1000G_NP1_UMAP_PC7_NC2_NN15_MD05_pca_hrshisp_added1kgp_2018115153245_admix.pdf}
    \end{subfigure}
    \caption{UMAP of the first 7 principal components of the Hispanic population of the HRS, coloured by estimated admixture proportions. Admixture proportions for each individual were estimated in \cite{baharian2016great} by assuming ancestral African, Asian, and European populations using RFMIX \cite{Maples:2013fia}. We have scaled each of the three proportions to values between 0 and 255 (with 100\% corresponding to 255), to colour individual points by their estimated admixture represented by RGB where red, green, and blue respectively correspond to African, European, and Asian/Native American ancestry.}
    \label{fig:supp_umap_hrs_hisp_admix}
\end{figure*}

@@

\begin{figure}
\centering
   \includegraphics[width=0.6\linewidth]{images/HRS_1000G_NP1_UMAP_PC10_NC2_NN15_MD05_pca_1kgp_onto_hrs_umap_1kgp_onto_hrs_2018112221116_admix.pdf}
   \caption{UMAP on the first 10 principal components of HRS data. colouring individuals by estimated admixture from three ancestral populations reveals considerable diversity in the Hispanic population. This projection coloured by self-identified race and Hispanic status is presented in figure~\ref{fig:supp_umap_hrs_eth}. Admixture proportions for each individual were estimated in \cite{baharian2016great} by assuming ancestral African, Asian, and European populations using RFMIX \cite{Maples:2013fia}. We have scaled each of the three proportions to values between 0 and 255 (with 100\% corresponding to 255), to colour individual points by their estimated admixture represented by RGB where red, green, and blue respectively correspond to African, European, and Asian/Native American ancestry.}
    \label{fig:umap_hrs_admix}
\end{figure}

@@

\begin{figure}[!htb]
    \centering
    \includegraphics[width=0.95\columnwidth]{images/1KGP_tsne_umap.jpeg}
    \caption{Comparing the visualizations of UMAP, standard t-SNE, and t-SNE initialized with a UMAP projection, on the top 10 principal components of the 1KGP. t-SNE used 5000 iterations. Initializing t-SNE with UMAP breaks the continuous structure of the projection and instead forms many small clusters.}
    \label{fig:supp_tsne_umap_compare_1kgp}
\end{figure}
